A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2025-04-30 manuscript.qmd
temporal

The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org). For each patient the primary tumor subsite is reportd and each indicidual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites.

@@

![Prevalence of ipsilateral LNL involvement stratified by subsite. The subsites are sorted in natural order to represent the continuously changing LNL involvement. The different tumo locations are color coded, where oral cavity subsistes are depicted in blue, larynx in green, hypopharynx in red and larynx in orange. The patient data is further stratified in early t-stage (0-2) and late t-stage (3-4). The legend furter specifies the number of patients in each subsite.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}

@@

::: {.callout-note}

## TODO: Discuss if we need this
<!-- Not sure we can skip it. We want to maximize this w.r.t. the model parameters. -->

@@

:::

@@

@@

By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the complete data likelihood (@eq-complete-likelihood).

>>>>

The datasets 1-4 are publicly available as CSV tables [@ludwig_multi-centric_2023;@ludwig_detailed_2022] and can be interactively explored on [LyProX](https://lyprox.org). For each patient the primary tumor subsite is reported and each individual LNL is reported as either metastatic or healthy given the available diagnostic modalities, which include pathology after neck dissection in some patients.
In this work we will stratify the tumor locations into different ICD codes which are depicted in @fig-subsites. For glottis tumors (C32.0), only t-stages T2,T3 and T4 have been taken into account as by definition glottis tumors are non-metastatic in the stages T0 and T1.

@@

![Prevalence of ipsilateral LNL involvement stratified by subsite. The subsites are sorted in natural order to represent the continuously changing LNL involvement. The different tumo locations are color coded, where oral cavity subsistes are depicted in blue, larynx in green, hypopharynx in red and larynx in orange. The patient data is further stratified in early t-stage (0-2) and late t-stage (3-4). For glottis patients t-stages 0 and 1 have been excluded from early t-stage. The legend furter specifies the number of patients in each subsite.](figures/involvement_I_to_V_all_sites_t_staging.png){#fig-involvement}

@@

@@

@@

It is important to note that this mixture model algorithm diverges from the conventional formulation of mixture models as introduced in @bishop_pattern_2006. In Bishop's work, the mixture model is defined with a single set of mixture parameters $\boldsymbol{\pi}$, implying that all data points are generated from the same mixture of models. In our approach, however, we categorize the data points (patients) into distinct cohorts (subsites), with each subsite characterized by its own set of mixture parameters. Consequently, we construct $S$ separate mixture models, each possessing different mixture parameters while all models share the same underlying probability distributions, specifically the varying lymphatic spread models.

@@

By iterating these steps, the EM algorithm is guaranteed to converge to a (local) maximum of the incomplete data likelihood (@eq-incomplete-likelihood).
